EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces

### 2018-08-28 sensors.tex
Added a new set of images. Completed math for every method.

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{EEG Waveform Analysis with applications to Brain Computer Interfaces}

@@

All these methods provide a feature that can be used as a template, whereas all of them are based on metrics that are extracted from the shape of the signal.  These features are used to create dictionaries or template databases.  Finally, these templates provide the basis for the pattern matching algorithm and offline classification.

@@

\begin{equation}
\varepsilon =  \left\lVert   x(n,c) - \sum_{i=1}^{m} w_i g_{yi}   \right\rVert
\label{eq:mperror}
\end{equation}

@@

\begin{equation}
\tilde{x_{0}}(n,c) = x(n,c), k \leftarrow 1
\label{eq:multiclassificationrow}
\end{equation}

@@

\begin{equation}
g_{yk}(n) = \arg \max_{g_{yi}} \left\lVert < \tilde{x_{k}}(n,c), g_{yi} > \right\rVert 
\label{eq:multiclassificationrow}
\end{equation}

@@

\begin{equation}
w_{k} =  \frac{< \tilde{x_{k}}(n,c), g_{yk} >}{g_{yk}}
\label{eq:multiclassificationrow}
\end{equation}

@@

\begin{equation}
\tilde{x_{k+1}}(n,c)=  \tilde{x_k}(n,c) - w_{k} g_{yk} (n), k \leftarrow k + 1
\label{eq:multiclassificationrow}
\end{equation}

Two variants of this algorithm are evaluated. In \textit{MP 1} the dictionary is directly constructed with the normalized templates from Fig \ref{fig:dictionaryfig}.  In \textit{MP 2} the coefficients of  Daubechies least-asymetric wavelet with 2 vanishing moments atoms are used to construct the dictionary.  The first version, the matching against the template is evaluated according to Equation \ref{eq:mperror} directly, whereas for the latter each feature is crafted by decomposing the signal in its coefficients and building a vector with them.

\begin{equation}
f =  {\bigg ( w_{i} \bigg )}_{1}^{D} 
\label{eq:multiclassificationrow}
\end{equation}

@@

Bond and Pompe Permutation Entropy has been extensively used in EEG processing, with applications on Anesthesia, Sleep Stage evaluation and particularly for Epilepsy pre-ictal detection \citep{Bandt2002}.  This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the number of times each sequence is found on the signal synthesized as numeric value called entropy~\citep{Nicolaou2010}. A signal on a window of length $W$ is represented by

@@

\begin{equation}
(x_n,x_{n+\tau},x_{n+2 \tau}...,x_{n+(m-1)\tau})
\label{eq:pesignal}
\end{equation}

@@

\begin{equation}
(\pi_{1},\pi_{2},...,\pi_{k})
\label{eq:pesignal}
\end{equation}

@@

\begin{equation}
P_i = \frac{1}{k} \sum_{s=1}^{k} \left[ \pi_{s} = \pi_{i} \right]
\label{eq:pesignal}
\end{equation}

@@

\begin{equation}
H(p) = \sum_{i=1}^{m!} p_{i} \; log \frac{1}{p_{i}}
\label{eq:pesignal}
\end{equation}

@@

\begin{equation}
f =  {\bigg ( H(p)_{u} \bigg )}_{1}^{N - (W + \tau (m - 1)) + 1}
\label{eq:multiclassificationrow}
\end{equation}

@@

This method is an extension of Slope Chain Code (SCC)~\citep{Alvarado-Gonzalez2016}. A coding scheme is generated from a sequence of sample points based on the angle between the horizontal line on a 2D-plane and any segment produced by two consecutive sample points, regarding them as coordinates on that plane.

@@

\begin{equation}
g = \lfloor n + ( m \Delta ) + 0.5 \rfloor
\label{eq:shccdelta}
\end{equation}

@@

\begin{equation}
\Delta = \bigg \lceil \frac{N}{G+1} \bigg \rceil
\label{eq:shccdelta}
\end{equation}

@@

\begin{equation}
m = sign \bigg (  \frac{N-1}{\Delta} \bigg )  \bigg \lfloor \left\lvert \frac{T-1}{\Delta} \right\lvert \bigg \rfloor
\label{eq:shccdelta}
\end{equation}

@@

\begin{equation}
e = \left[ (x',y')_{1}, (x',y')_{s}, ..., (x',y')_{G} \right]
\label{eq:shccdelta}
\end{equation}

@@

\begin{equation}
\textbf{x} = \frac{1}{\max(x') - min(x')} ( \textbf{x'} - \min(x') \textbf{1} )
\label{eq:shccdelta}
\end{equation}

\begin{equation}
\textbf{y} = \frac{1}{\max(y') - min(y')} ( \textbf{y'} - \min(y') \textbf{1} )
\label{eq:shccdelta}
\end{equation}

@@

\begin{equation}
f = \bigg (  \frac{y_{s}-y_{s-1}}{x_{s}-x_{s-1}}  \bigg )_{2}^{G} 
\label{eq:shccdelta}
\end{equation}

\subsection{Histogram of Gradients Orientations - HGO}

This Histogram of Gradient Orientations is based on using Computer Vision techniques to extract directly information from a signal based on the plot of a signal on a 2D image.  This mimicks exactly what a person is doing by visually inspecting the plot, and the waveform.  To do that, the region of the image where the signal is located is divided in a $4 x 4$ block and the signal's bidimensional gradients is calculated.  For each block within the patch, a histogram of gradient orientations, 8 circular orientations, is calculated.  This histogram is concatenated for all the 16 blocks and a feature is thus formed.  The details of the method can be found here \citep{Ramele2016}.  This method was implemented using the VLFeat  \citep{Vedaldi2010} Computer Vision libraries.

\begin{equation}
f = {\bigg ( h(i,j,\theta) \bigg )}_{i,j,\theta}
\label{eq:multiclassificationrow}
\end{equation}

@@

Farwell and Donchin P300 Speller \citep{Farwell1988} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing in a virtual keyboard.  This procedure exploits a cognitive phenomena called oddball paradigm: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.  This is cleverly utilized in the P300 Speller, where rows and columns of a 6x6 matrix flashes randomly but only the flashing of a column or row where the letter that a user is focusing will trigger concurrently the P300 ERP along the EEG trace.

%Algorithms based on qEEG can detect this component in a good-enough single-trial approach [P300 Summary], but at the same time, these algorithms do not establish a clear relationship with the P300 ERP. Hence it is difficult to establish a backward mapping back to a certain visual pattern within the waveform.

In order to verify each algorithm in a controlled procedure, a real ERP template is superimposed into a real EEG stream.  This trace was experimentally obtained by a subject which was observing the flashing of the stimulus matrix during a P300 Speller procedure but did not engage in focusing on anyone in particular. Everything is there, except the P300 ERP component. By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential in accordance to a similar procedure used in similar works \citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.

@@

\begin{itemize}
\item Pseudo-real dataset Classification Performance: the letter classification performance of each one of these methods.
\item Latency Noise: Gaussian noise is added to the latency where the ERP template is injected into the EEG stream.
\item Component Amplitude Noise: the different components of the ERP template are altered to make them less distinguishable.
\end{itemize}

Finally results for the same methods while offline processing a public dataset of real P300 patterns are also shown.

@@

The template ERP is extracted from the Subject $8$ of the public dataset 008-2014  \citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia. Segments from the EEG signal are labeled as hit and are extracted for the trial number $2$, and they are point-to-point coherently averaged.  This P300 ERP can be seen in Fig.  \ref{fig:erptemplate1}.

An EEG stream with null-P300 signal everywhere is obtained by the following procedure: 
The participant is recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.  No monetary compensation is handed out and he/she agree and sign a written informed consent.  This study is approved by the \textit{Departamento de Investigación y Doctorado, Instituto Tecnológico de Buenos Aires (ITBA)}.  The participant is healthy and have normal or corrected-to-normal vision and no history of neurological disorders. This voluntary is of unspecified gender, aged between 20-30 years old.  EEG data is collected in a single recording session. She/He is seated in a comfortable chair, with her/his vision aligned to a computer screen located one meter in front of him.  The handling and processing of the data and stimuli is conducted by the OpenVibe platform~\citep{Renard2010}.  Gel-based active electrodes (g.LADYbird, g.Tec, Austria) are used on locations Fz, Cz, Pz, Oz, P3,P4, PO7 and PO8 according to the 10-20 international system.  Reference is set to the right ear lobe and ground is preset as the AFz position.   Sampling frequency is set to 250 Hz, which is the closest possible to the one used with the other dataset.

@@

\begin{figure}[H]
\centering
\includegraphics[width=10cm]{images/doublegain.eps}
\includegraphics[width=10cm]{images/triplegain.eps}
\caption{Eight-channel EEG signal superimposed with the ERP Template.  (Left) EEG trace of the original signal. (Right) The same signal segment with the added template.}
\label{fig:doubleandtriplegain}
\end{figure}

>>>>

\usepackage{subfigure}

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces}

@@

All these methods provide a feature $ f $  that can be used as a template, whereas all of them are based on metrics that are extracted from the shape of the single channel digital EEG signal $x(n)$, with $n$ varying from $1$ to the length of the EEG segment in sample points.  These features are used to create dictionaries or template databases.  Finally, these templates provide the basis for the pattern matching algorithm and offline classification.

@@

\begin{equation}
\varepsilon =  \left\lVert   x(n) - \sum_{i=1}^{m} w_i g_{yi}(n)   \right\rVert
\label{eq:mperror}
\end{equation}

@@

\begin{equation}
\tilde{x_{0}}(n) = x(n), k \leftarrow 1
\label{eq:mp2}
\end{equation}

@@

\begin{equation}
g_{yk}(n) = \arg \max_{g_{yi}} \left\lVert < \tilde{x_{k}}(n), g_{yi}(n) > \right\rVert 
\label{eq:mp3}
\end{equation}

@@

\begin{equation}
w_{k} =  \frac{< \tilde{x_{k}}(n), g_{yk}(n) >}{  | g_{yk} | }
\label{eq:mp4}
\end{equation}

@@

\begin{equation}
\tilde{x_{k+1}}(n)=  \tilde{x_k}(n) - w_{k} g_{yk} (n), k \leftarrow k + 1
\label{eq:mp5}
\end{equation}

Two variants of this algorithm are evaluated. In \textit{MP 1} the dictionary is directly constructed with the normalized templates from Fig \ref{fig:dictionaryfig}.  In \textit{MP 2} the coefficients of  Daubechies least-asymetric wavelet with 2 vanishing moments atoms are used to construct the dictionary.  For the first version, the matching against the template is evaluated according to Equation \ref{eq:mperror} directly, whereas for the latter each feature is crafted by decomposing the signal in its coefficients and building, an eventually sparse, vector with them.

\begin{equation}
f =  {\bigg ( w_{i} \bigg )}_{1}^{D} 
\label{eq:mp6}
\end{equation}

@@

Bond and Pompe Permutation Entropy has been extensively used in EEG processing, with applications on Anesthesia, Sleep Stage evaluation and significantly for Epilepsy pre-ictal detection \citep{Bandt2002}.  This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the number of times each sequence is found on the signal synthesized as a numeric value called entropy~\citep{Nicolaou2010}. A signal on a window of length $W$ is represented by

@@

\begin{equation}
(x_n,x_{n+\tau},x_{n+2 \tau}...,x_{n+(m-1)\tau})
\label{eq:pe2}
\end{equation}

@@

\begin{equation}
(\pi_{1},\pi_{2},...,\pi_{k})
\label{eq:pe3}
\end{equation}

@@

\begin{equation}
P_i = \frac{1}{k} \sum_{s=1}^{k} \left[ \pi_{s} = \pi_{i} \right]
\label{eq:pe4}
\end{equation}

@@

\begin{equation}
H(p) = \sum_{i=1}^{m!} p_{i} \; log \frac{1}{p_{i}}
\label{eq:pe5}
\end{equation}

@@

\begin{equation}
f =  {\bigg ( H(p)_{u} \bigg )}_{1}^{N - (W + \tau (m - 1)) + 1}
\label{eq:pe6}
\end{equation}

@@

This method is an extension of Slope Chain Code (SCC)~\citep{Alvarado-Gonzalez2016}. This algorithm proceeds by generating a coding scheme from a sequence of sample points. This encoding is based on the angle between the horizontal line on a 2D-plane and any segment produced by two consecutive sample points, regarding them as coordinates on that plane.

@@

\begin{equation}
g = \lfloor n + ( m \Delta ) + 0.5 \rfloor
\label{eq:shcc2}
\end{equation}

@@

\begin{equation}
\Delta = \bigg \lceil \frac{N}{G+1} \bigg \rceil
\label{eq:shcc3}
\end{equation}

@@

\begin{equation}
m = sign \bigg (  \frac{N-1}{\Delta} \bigg )  \bigg \lfloor \left\lvert \frac{T-1}{\Delta} \right\lvert \bigg \rfloor
\label{eq:shcc4}
\end{equation}

@@

\begin{equation}
e = \left[ (x',y')_{1}, (x',y')_{s}, ..., (x',y')_{G} \right]
\label{eq:shcc5}
\end{equation}

@@

\begin{equation}
\textbf{x} = \frac{1}{\max(x') - min(x')} ( \textbf{x'} - \min(x') \textbf{1} )
\label{eq:shcc6}
\end{equation}

\begin{equation}
\textbf{y} = \frac{1}{\max(y') - min(y')} ( \textbf{y'} - \min(y') \textbf{1} )
\label{eq:shcc7}
\end{equation}

@@

\begin{equation}
f = \bigg (  \frac{y_{s}-y_{s-1}}{x_{s}-x_{s-1}}  \bigg )_{2}^{G} 
\label{eq:shcc8}
\end{equation}

\subsection{Scale Invariant Feature Transform - SIFT}

SIFT~\citep{Lowe2004} is a very successful feature extraction technique from Computer Vision .  It has a biomimetic inspiration on how the visual cortex analyze images based on orientations.  This method has been used in \citep{Ramele2016} to analyze EEG signals based on their plots on 2D images.

The first step of the algorithm is plot generation based on single-channel EEG segments $ x(n) $.

\begin{equation}
\tilde{x}(n) = \left\lfloor \frac{\delta ( x(n) - \bar{x})}{\sigma_{x}} \right\rceil 
\label{eq:sift1}
\end{equation}

\noindent with $\delta$ being the signal magnification factor and $\bar{x}$ and $\sigma_{x}$ the mean and standard deviation of $x$ on the signal segment.  The width of the image is determined straightforward based on the $1$-second length in sample units. This corresponds to the sampling frequency $F_s$ of the EEG signal segment. The width is adjusted by multiplying by a magnification factor $\delta$,

\begin{equation}
w = \delta \; F_s
\label{eq:sift2}
\end{equation}

\noindent whereas the height is calculated based on the peak-to-peak amplitude of the signal within the segment,

\begin{equation}
h = ( \max_{n} \tilde{x}(n) - \min_{n} \tilde{x}(n))
\label{eq:sift4}
\end{equation}

Equation \ref{eq:sift5} determines the height of the image where the signal's zero value will be located.

\begin{equation}
z = \left\lfloor \frac{h}{2} \right\rfloor - \left\lfloor \frac{\max_{n} \tilde{x}(n) + \min_{n} \tilde{x}(n)}{2} \right\rfloor
\label{eq:sift5}
\end{equation}

Finally, a binary, black-and-white image plot is generated based on

\begin{equation}
I(z_1,z_2) = \left\{ \begin{array}{rl}
255 & \text{if} \,  z_1 = \delta \  n; \! z_2 = \tilde{x}(n) + z \\
0   & \mbox{otherwise}
\end{array}\right.
\label{eq:sift6}
\end{equation}

Once the plot is generated, the plot's center is used to localize the SIFT patch. This region of the image, where the signal's most important salient waveform should be located, is divided in a grid of $4 x 4$ block and the bidimensional gradient vectors are calculated on each one of them.  At the same time, for each block within the patch, a histogram of the gradient orientations, 8 circular orientations, are calculated.  This histogram is concatenated for all the 16 blocks and a feature is thus formed:

\begin{equation}
f = {\bigg ( h(i,j,\theta) \bigg )}_{i,j,\theta}
\label{eq:sift7}
\end{equation}

\noindent with $i$ and $j$ varying between $0$ and $3$, localizing the 16 blocks within the grid. The angles $\theta$ are the eight possible values between $0$ and $315$.  This vector is normalized, clapped to $0.2$, and re-normalized again.   Details of the method can be found on \citep{Ramele2016,Lowe2004}.  It was implemented using the VLFeat  \citep{Vedaldi2010} public Computer Vision libraries.

@@

Farwell and Donchin P300 Speller \citep{Farwell1988} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing in a virtual keyboard.  This procedure exploits a cognitive phenomena raised by the oddball paradigm: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.  This is cleverly utilized in the P300 Speller, where rows and columns of a 6x6 matrix flashes randomly but only the flashing of a column or row where the letter that a user is focusing will trigger concurrently the P300 ERP along the EEG trace.

How much is it possible to classify and implement a valid P300 ERP Speller using only the waveform structure of the signal and which algorithm perform better at this even when single trial latency variations and component amplitudes issues are present ?
In order to verify this in a controlled procedure, a real P300 ERP template, obtained from a public dataset, is superimposed into a different EEG stream.  This trace was experimentally obtained by a subject which was observing the flashing of the stimulus matrix during a P300 Speller procedure but did not engage in focusing on anyone in particular. Everything is there, except the P300 ERP component. By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential in accordance to a similar procedure used in similar works \citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.

@@

\begin{itemize}
\item Pseudo-real dataset Classification Performance: the letter classification performance of each one of these methods.
\item Latency Noise: Gaussian noise is added to the ERP template latency, i.e. the temporal location where it is injected into the EEG stream.
\item Component Amplitude Noise: the different components of the ERP template are altered to make them less distinguishable.
\end{itemize}

\textbf{Borrar o hacer: Finally results for the same methods while offline processing a public dataset of real P300 patterns are also shown.}

@@

The template ERP is extracted from the Subject Number $8$ of the public dataset 008-2014  \citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia. Segments from the EEG signal are labeled as hit and are extracted for the trial number $2$, and they are point-to-point coherently averaged.  This P300 ERP can be seen in Fig.  \ref{fig:erptemplate1}.

An EEG stream with null-P300 signal is obtained by the following procedure: 
The participant is recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.  No monetary compensation is handed out and he/she agree and sign a written informed consent.  This study is approved by the \textit{Departamento de Investigación y Doctorado, Instituto Tecnológico de Buenos Aires (ITBA)}.  The participant is healthy and have normal or corrected-to-normal vision and no history of neurological disorders. This voluntary is of unspecified gender, aged between 20-30 years old.  EEG data is collected in a single recording session. She/He is seated in a comfortable chair, with her/his vision aligned to a computer screen located one meter in front of him.  The handling and processing of the data and stimuli is conducted by the OpenVibe platform~\citep{Renard2010}.  Gel-based active electrodes (g.LADYbird, g.Tec, Austria) are used on locations Fz, Cz, Pz, Oz, P3,P4, PO7 and PO8 according to the 10-20 international system.  Reference is set to the right ear lobe and ground is preset as the AFz position.   Sampling frequency is set to 250 Hz, which is the closest possible to the one used with the other dataset.

@@

\begin{figure}[htb]
\centering
\subfigure[EEG trace of the original signal.]{\includegraphics[width=.45\linewidth]{images/nogain.eps}}
\subfigure[The same signal segment with the added template.]{\includegraphics[width=.45\linewidth]{images/singlegain.eps}}
\subfigure[EEG sample of Cz channel of the original EEG trace.]{\includegraphics[width=.45\linewidth]{images/nogainzoomhit.eps}}
\subfigure[The same segment with the added template.]{\includegraphics[width=.45\linewidth]{images/singlegainzoomhit.eps}}
\caption{Eight-channel EEG signal superimposed with the ERP Template.}
\label{fig:doubleandtriplegain}
\end{figure}

%\begin{figure}[H]
%\centering
%\includegraphics[width=12cm]{images/nogain.eps}
%\includegraphics[width=12cm]{images/singlegain.eps}
%\caption{Eight-channel EEG signal superimposed with the ERP Template.  (Left) EEG trace of the original signal. (Right) The same signal segment with the added template.}
%\label{fig:doubleandtriplegain}
%\end{figure}
